---
title: "Tutorial"
---

# Overview

The necessary code for the workshop will be hosted here, alongside questions and instructions that build on the code posted

You can copy and paste the code here into your file

You can answer questions as comments in the R chunks

# Section 1: R basics Tutorial

```{r}
#| results: hide    

#R Concepts Tutorial - Part 1 

#Focus on this tutorial will not be to necessarily understand the basics of R programming, but more to give you the tools to understand the workflow 

# You can copy and paste the code found here into your own files and run the code yourself 


# 1) By putting a hastag in front of text you can put a "comment" 

# R will not interpret this as code

# We can use this as a way to annotate our code with helpful notes 




# 2) We will be using a R markdown file to write our code

# It allows us organize our code by keeping sections in different "chunks"


# You can create a new chunk by clicking "Insert a new code chunk" in the top bar (button with a green C with a plus)





```

```{r}

# R Concepts - Part 2


#In R, you can save information in an object called a variable. This lets you reuse a value, data frame, or result later without having to type it again.

#Use `<-` to assign something to a variable:


# Save the number 5 in a variable named "number_of_samples"

# You can run this line of code by higlighting the line and selecting "run selected lines" on the "Run" button on the top bar

#Alternatively, you can run all lines in a chunk by selecting "Run Current Chunk" or By pressing the Green button at the top right of the chunk


number_of_samples <- 5


#you can view the number_of_samples variable by viewing it in the "environment" tab or by running just the variable 
number_of_samples


```

```{r}
# R concepts Part 3 

# R Concepts Tutorial - Part 3: Functions

# A function is a reusable command that performs a specific task.

# Functions usually have:
# 1. A function name
# 2. Parentheses ()
# 3. An input inside the parentheses, called an "argument"

# For example, the c() function combines several values into one vector (ordered collection of values):

sample_ids <- c("Sample_01", "Sample_02", "Sample_03")


# Type the variable name and run the line to view its contents:

sample_ids



# The length() function counts the number of values in an object:

length(sample_ids)



# If you are unsure what a certain function does, and what argument it takes, you can put a "?" before the function and run that to bring up the documentation 

?length

```

```{r}
#| results: hide    
#| message: false
#| warning: false

# R concepts Part 4

#Certain functions that we may need are not automatically loaded in R 
#These functions are found within R "packages" that contain a set of functions 
#To load these packages, we can use the "library" function 

#For our workshop  we will need the following: 

library(dplyr) #manipulating dataframes etc.
library(phyloseq) #integrates taxa, abundance, sample data into integrated object for analyses
library(microbiome) #performs functions on phyloseq object to transform/visualize
library(microViz)  #helps perform visualizations on phyloseq objects 
library(vegan) #ecological distance calculations/stats
library(stringr) #working with character strings


```

```{r}
#| results: hide    
#| message: false
#| warning: false
# R concepts Part 5 

# We can load our phyloseq object and save it as the "phyloseq.dogs" variable by using the readRDS function and using the pathname where the object is found as an argument

#Replace the pathfile to where your file is located 
phyloseq.dogs <-  readRDS("/Users/anshulsinha/Desktop/microbiome_center/centre-symposium-2026/R_material/dogs_phyloseq.rds")

#If using posit.cloud, run the following: 

#phyloseq.dogs <- readRDS("/cloud/project/R_material/dogs_phyloseq.rds")

```

# Section 2: Characterizing the Dataset and Phyloseq Object

Now that we have our phyloseq object, let's survey what we have here

```{r}
#| results: hide    
#| message: false
#| warning: false
#run the phyloseq object to get basic characteristics of the dataset

phyloseq.dogs

```

```{r}
#| results: hide    
#| message: false
#| warning: false

#you can run the phyloseq object it's chunk to return some basic characteristics of the dataset 


phyloseq.dogs



```

How many unique species are in the dataset?

How many samples?

How many metadata categories do we have?

Let's look at each component of the phyloseq object

```{r}
#| results: hide    
#1) The OTU_table function can acess the "feature table"
# We can save the feature table as a data frame (a table-like data structure)
# Note: here we are using a function (otu_table) within a function (as.data.frame)

phyloseq.dogs.otu.table <- as.data.frame(otu_table(phyloseq.dogs))

#remember, you can view the variables by either running the variable itself as a line of code, or looking for it in the environment panel 



```

Any immediate observations here?

```{r}
#| results: hide    
#2) The tax_table function allows us to access the "taxonomy table"
phyloseq.dogs.tax.table <- as.data.frame(tax_table(phyloseq.dogs))


# Each row of the tax.table is a unique species that is found at least once in the dataset

# You can inspect how the unique phyla by doing the following 
# The "$" operator accesses a column of the data.frame 
unique(phyloseq.dogs.tax.table$phylum)


#you can add the "length" function to give the number of unique phyla 
length(unique(phyloseq.dogs.tax.table$phylum))

```

How many unique family are there?

genera?

```{r}
#| results: hide    
#3) The sample_data function allows us to access the "metadata"
phyloseq.dogs.metadata <- as.data.frame(sample_data(phyloseq.dogs))

#Go through the metadata columns to see what sorts of information we have on the samples 
# Remember, this dataset contains dogs sampled longitudinally 

#Start to think about what might categories might be interesting to test their associations with the microbiome



```

How many unique dogs do we have?

diet groups?

```{r}
#| results: hide    
#We can always subset our phyloseq object to work with a subset of samples if we need to 

# We can do this by using the subset_samples function a given metadata category 

# Let's get a phyloseq object of just puppy samples (indicated by the "host_life_stage column)


phyloseq.dogs.puppies <- subset_samples(phyloseq.dogs, host_life_stage == "puppy")

```

how many puppy samples are there?

# Section 3: Visualization of Microbiome Data using Bar Plots

-   Now that we have a sense of the samples we are working with, lets try to visualize some data

-   A good way to get a first overview of microbiome data is through with simple bar plots. These aren’t perfect, but they can give us a nice first impression of what to expect

-   We can make a simple bar plot using the Microviz comp_barplot command

-   This command will plot (per sample) the relative abundance of bacteria at different taxonomic levels

```{r}
# We will be using a command called "comp_barplot" which takes our phyloseq object and creates relative abundance plots 

#step 1) Use the tax_fix command to make the taxonomy appropriate for the comp_barplot function 

#Lets do this using our phyloseq object that we originally made 

fixed.physeq <- tax_fix(phyloseq.dogs)


#Step 2) We can use the comp_barplot to create a barplot 
# It takes a minimum of two arguments, each separated by a comma 1) The "fixed" phyloseq object and 2) the taxonomic rank at which to create the plots 

#we also added the "label = NULL" argument to not plot sample names to keep our plot cleaner 



comp_barplot(fixed.physeq, tax_level = "phylum", label = NULL)


#remember that we can run the following to look at the documentation of the function 
?comp_barplot
  
#Not all these will be relevant, but some, like n_taxa can control how many distinct taxa we plot 

comp_barplot(fixed.physeq, tax_level = "phylum", n_taxa = 12, label = NULL) 

```

What are the most abundant phyla in the dataset?

Go through different levels of taxonomy, and report the most abundant members of the community.

What levels of variation do you see between samples?

Does this variation change as the levels of taxonomy become more resolute?

```{r}
#| results: hide    

#It is probably useful to try to see how our taxa composition differs by different metadata categories

#We can use the "group_by" argument to split our plots by a given category 


#Remember, we can look at our metadata table by doing the following: 
sample_data(fixed.physeq)


#For instance to look at how composition differs between puppies and adults at the family level, we can do the following

comp_barplot(fixed.physeq, tax_level = "family", n_taxa = 12, label = NULL, group_by = "host_life_stage") 


```

Are there obvious differences between the composition of the adult and puppies?

If so, describe some general trends

Does one group have more variation than the other?

```{r}
#| results: hide    

#Let's focus just on the puppies
#we can use the subsetted phyloseq object that we created earlier

phyloseq.dogs.puppies <- subset_samples(phyloseq.dogs, host_life_stage == "puppy")


#Reminder: we have to run tax_fix on this first 
fixed.physeq.puppy <- tax_fix(phyloseq.dogs.puppies)


#Then, we can run comp_barplot like we did earlier and test the effect of different metadata categories 

#Remember, you can run the following to look at the metadata

phyloseq.dogs.metadata.puppies <- as.data.frame(sample_data(phyloseq.dogs.puppies))




```

Are there metadata categories that seem to influence puppy microbiome composition more than others?

Try to identify some species that are associated with a given group

Since this is a longitudinal dataset, it may be interesting to follow the composition of a single dog over time. See if you can do this with the dog id "DMD_E_Rozsaszin". Note: you'll need to use the "host_subject_id" metadata column

# Optional Section 3.1: Plotting individual taxa

```{r}
#Now that we've identified some taxa of interest, we can try plotting them individually to get a more clear look at their dynamics 

#Let's plot how E. coli changes over age group 

#We can use the tax_transform function to change our phyloseq object and normalize it using relative abundance and get our phyloseq object at the species level

#it takes the phyloseq object as an argument, the "compositional" string to indicate we are transforming to get a relative abundance phyloseq object, and rank to do this at the genus level 

phyloseq.dogs.puppies.relab.genus <- tax_transform(fixed.physeq.puppy, "compositional", rank = "genus") #


# lets turn this back into a phyloseq object using the ps_get function

phyloseq.dogs.puppies.relab.genus.ps <- ps_get(phyloseq.dogs.puppies.relab.genus) #returns a phyloseq object 


#now we can add the relative abundance of Escherichia to the sample data as a column using ps_otu2samdat

phyloseq.dogs.puppies.relab.genus.ps.escherichia <- ps_otu2samdat(phyloseq.dogs.puppies.relab.genus.ps, "Escherichia") 


#now we can make a table with its abundance 
phyloseq.dogs.puppies.relab.genus.ps.escherichia.table <-   samdat_tbl(phyloseq.dogs.puppies.relab.genus.ps.escherichia)

#You can see there is a column called Escherichia with the relative abundance of Escherichia in every sample 



#Now we can use ggplot2 to plot the abundance of this species 
#ggplot2 takes our table as an argument, and then asks for our x and y-axes in the aes function 

# we can specify we want this to be a boxplot using geom_boxplot
# we can specify that we want to add individual points with geom jitter
ggplot(phyloseq.dogs.puppies.relab.genus.ps.escherichia.table
, aes(x = host_age_group, y = Escherichia, colour = host_age_group)) +
  geom_boxplot(width = 0.5, colour = "grey") +
  geom_jitter(width = 0.2, alpha = 0.5) 



```

Give this a try with another bacterial genus!

# Section 4: Normalizing Sequencing Depth using Rarefaction

-   While our bar plots have given us some nice qualitative measures of the community composition, can we assess changes in a more quantitative way?

-   We can calculate changes in microbial diversity (the number of taxa present and how they are distributed)

-   As covered in the lectures, we can assess within-sample diversity (alpha diversity) and between-sample diversity (beta-diversity). This involves calculating different diversity metrics and calculating ecological distance

-   Before we do this, there is an important consideration: library size normalization!

-   During a sequencing run, different samples will inevitably have different numbers of reads, which means we have to normalize the read depth of our samples to make accurate comparisons

-   For the bar plots, we’ve calculated relative abundance, which essentially “normalized” the samples (reads of taxa/total reads in the sample) to make them comparable. Using relative abundance for normalization in diversity calculations is not generally advised. There is a lot of debate and controversy around what normalization approaches are the best.

-   In reality, no single method is perfect. But we will be using a method called rarefaction. It’s relatively simple and a recently published [paper](https://pmc.ncbi.nlm.nih.gov/articles/PMC10900887/) by Schloss put forward that it is the best method for normalizing amplicon sequencing data

-   Basically, rarefaction involves the random subsampling of reads in each sample to the SAME depth

-   To illustrate the concept of rarefaction lets look at the distribution of reads in our samples

```{r}
#Above, we used total sum scaling to look at the proportions of bacterial taxa

#As a technical artifact of sequencing, we may have different numbers of reads between samples 
#We can look at the number of reads in each sample by using the "sample_sums" function on our phyloseq object 

sample_sums(phyloseq.dogs.puppies)

#we can find the min and max number of reads in our phyloseq object by doing the following
min(sample_sums(phyloseq.dogs.puppies))
max(sample_sums(phyloseq.dogs.puppies))


#we can visualize this by saving the sample_sums as a dataframe and then plotting as a histogram 

#save sample_sums as the variable "lib_sums"


lib_sizes <- sample_sums(phyloseq.dogs.puppies)

#can visualize lib_sizes using the "hist" function

hist(lib_sizes,breaks = 50)



hist(
  lib_sizes,
  breaks = 30,
  xlab = "Total reads per sample",
  col = "lightblue",
  border = "white"
)


```

Now, let's build a rarefaction curve to see if we've sequenced enough

```{r}
#First let's get a matrix from our phyloseq object
#We are using "t" to transpose the rows and columns (rarecurve takes the taxa and abundance in opposite oritentation to phyloseq)

matrix.phyloseq.dogs.puppies <- as(t(otu_table(phyloseq.dogs.puppies)), "matrix")

## Calculate row sums (sequencing depth of each sample)

row.sums.puppies <- rowSums(matrix.phyloseq.dogs.puppies)


#Our curve might be messy if we plot ALL of our samples 

# Lets order rows by row sums and take the 50 rows with the lowest sums
matrix.phyloseq.dogs.puppies.low50 <-
  matrix.phyloseq.dogs.puppies[order(row.sums.puppies)[1:50], ]


?rarecurve
#now we have a matrix that we can put into the rarecurve function! 
#step = 1000  means that the function calculates species richness every 1000 sequences until the total number of species for that sample is reached

rarecurve(matrix.phyloseq.dogs.puppies.low50, step = 1000, label = FALSE)


```

Are all samples plotted sequenced to sufficient depth?

What would be a good depth to rarefy at?

```{r}
#Our lowest sequencing depth is: 
min(sample_sums(phyloseq.dogs.puppies))


```
