---
title: "Tutorial"
---

# Overview

The necessary code for the workshop will be hosted here, alongside questions and instructions that build on the code posted

You can copy and paste the code here into your file

# Section 1: R basics Tutorial

```{r}
#| eval: false

#R Concepts Tutorial - Part 1 

#Focus on this tutorial will not be to necessarily understand the basics of R programming, but more to give you the tools to understand the workflow 

# You can copy and paste the code found here into your own files and run the code yourself 


# 1) By putting a hastag in front of text you can put a "comment" 

# R will not interpret this as code

# We can use this as a way to annotate our code with helpful notes 




# 2) We will be using a R markdown file to write our code

# It allows us organize our code by keeping sections in different "chunks"


# You can create a new chunk by clicking "Insert a new code chunk" in the top bar (button with a green C with a plus)





```

```{r}
#| eval: false

# R Concepts - Part 2


#In R, you can save information in an object called a variable. This lets you reuse a value, data frame, or result later without having to type it again.

#Use `<-` to assign something to a variable:


# Save the number 5 in a variable named "number_of_samples"

# You can run this line of code by higlighting the line and selecting "run selected lines" on the "Run" button on the top bar

#Alternatively, you can run all lines in a chunk by selecting "Run Current Chunk" or By pressing the Green button at the top right of the chunk


number_of_samples <- 5


#you can view the number_of_samples variable by viewing it in the "environment" tab or by running just the variable 
number_of_samples


```

```{r}
#| eval: false

# R concepts Part 3 

# R Concepts Tutorial - Part 3: Functions

# A function is a reusable command that performs a specific task.

# Functions usually have:
# 1. A function name
# 2. Parentheses ()
# 3. An input inside the parentheses, called an "argument"

# For example, the c() function combines several values into one vector:

sample_ids <- c("Sample_01", "Sample_02", "Sample_03")


# Type the variable name and run the line to view its contents:

sample_ids



# The length() function counts the number of values in an object:

length(sample_ids)


```

```{r}
#| eval: false

# R concepts Part 4

#Certain functions that we may need are not automatically loaded in R 
#These functions are found within R "packages" that contain a set of functions 
#To load these packages, we can use the "library" function 

#For our workshop  we will need the following: 

library(dplyr) #manipulating dataframes etc.
library(phyloseq) #integrates taxa, abundance, sample data into integrated object for analyses
library(microbiome) #performs functions on phyloseq object to transform/visualize
library(microViz)  #helps perform visualizations on phyloseq objects 
library(vegan) #ecological distance calculations/stats
library(stringr) #working with character strings


```

```{r}
#| eval: false

# R concepts Part 5 

# We can load our phyloseq object and save it as the "phyloseq.dogs" variable by using the readRDS function and using the pathname where the object is found as an argument

#Replace the pathfile to where your file is located 
phyloseq.dogs <-  readRDS("/Users/anshulsinha/Desktop/microbiome_center/centre-symposium-2026/R_material/dogs_phyloseq.rds")

#If using posit.cloud, run the following: 

#phyloseq.dogs <- readRDS("/cloud/project/R_material/dogs_phyloseq.rds")

```

# Section 2: Characterizing the Dataset and Phyloseq Object

Now that we have our phyloseq object, let's survey what we have here

```{r}
#| eval: false

#run the phyloseq object to get basic characteristics of the dataset

phyloseq.dogs

```

How many samples are in the dataset?

species?

metadata categories?
