---
title: "Tutorial"
---

# Overview

The necessary code for the workshop will be hosted here, alongside questions and instructions that build on the code posted

You can copy and paste the code here into your file

You can answer questions as comments in the R chunks

# Section 1: R basics Tutorial

```{r}
#| results: hide    

#R Concepts Tutorial - Part 1 

#Focus on this tutorial will not be to necessarily understand the basics of R programming, but more to give you the tools to understand the workflow 

# You can copy and paste the code found here into your own files and run the code yourself 


# 1) By putting a hastag in front of text you can put a "comment" 

# R will not interpret this as code

# We can use this as a way to annotate our code with helpful notes 




# 2) We will be using a R markdown file to write our code

# It allows us organize our code by keeping sections in different "chunks"


# You can create a new chunk by clicking "Insert a new code chunk" in the top bar (button with a green C with a plus)





```

```{r}

# R Concepts - Part 2


#In R, you can save information in an object called a variable. This lets you reuse a value, data frame, or result later without having to type it again.

#Use `<-` to assign something to a variable:


# Save the number 5 in a variable named "number_of_samples"

# You can run this line of code by higlighting the line and selecting "run selected lines" on the "Run" button on the top bar

#Alternatively, you can run all lines in a chunk by selecting "Run Current Chunk" or By pressing the Green button at the top right of the chunk


number_of_samples <- 5


#you can view the number_of_samples variable by viewing it in the "environment" tab or by running just the variable 
number_of_samples


```

```{r}
# R concepts Part 3 

# R Concepts Tutorial - Part 3: Functions

# A function is a reusable command that performs a specific task.

# Functions usually have:
# 1. A function name
# 2. Parentheses ()
# 3. An input inside the parentheses, called an "argument"

# For example, the c() function combines several values into one vector (ordered collection of values):

sample_ids <- c("Sample_01", "Sample_02", "Sample_03")


# Type the variable name and run the line to view its contents:

sample_ids



# The length() function counts the number of values in an object:

length(sample_ids)



# If you are unsure what a certain function does, and what argument it takes, you can put a "?" before the function and run that to bring up the documentation 

?length

```

```{r}
#| results: hide    
#| message: false
#| warning: false

# R concepts Part 4

#Certain functions that we may need are not automatically loaded in R 
#These functions are found within R "packages" that contain a set of functions 
#To load these packages, we can use the "library" function 

#For our workshop  we will need the following: 

library(dplyr) #manipulating dataframes etc.
library(phyloseq) #integrates taxa, abundance, sample data into integrated object for analyses
library(microbiome) #performs functions on phyloseq object to transform/visualize
library(microViz)  #helps perform visualizations on phyloseq objects 
library(vegan) #ecological distance calculations/stats
library(stringr) #working with character strings


```

```{r}
#| results: hide    
#| message: false
#| warning: false
# R concepts Part 5 

# We can load our phyloseq object and save it as the "phyloseq.dogs" variable by using the readRDS function and using the pathname where the object is found as an argument

#Replace the pathfile to where your file is located 
phyloseq.dogs <-  readRDS("/Users/anshulsinha/Desktop/microbiome_center/centre-symposium-2026/R_material/dogs_phyloseq.rds")

#If using posit.cloud, run the following: 

#phyloseq.dogs <- readRDS("/cloud/project/R_material/dogs_phyloseq.rds")

```

# Section 2: Characterizing the Dataset and Phyloseq Object

Now that we have our phyloseq object, let's survey what we have here

```{r}
#| results: hide    
#| message: false
#| warning: false
#run the phyloseq object to get basic characteristics of the dataset

phyloseq.dogs

```

```{r}
#| results: hide    
#| message: false
#| warning: false

#you can run the phyloseq object it's chunk to return some basic characteristics of the dataset 


phyloseq.dogs



```

How many unique species are in the dataset?

How many samples?

How many metadata categories do we have?

Let's look at each component of the phyloseq object

```{r}
#| results: hide    
#1) The OTU_table function can acess the "feature table"
# We can save the feature table as a data frame (a table-like data structure)
# Note: here we are using a function (otu_table) within a function (as.data.frame)

phyloseq.dogs.otu.table <- as.data.frame(otu_table(phyloseq.dogs))

#remember, you can view the variables by either running the variable itself as a line of code, or looking for it in the environment panel 



```

Any immediate observations here?

```{r}
#| results: hide    
#2) The tax_table function allows us to access the "taxonomy table"
phyloseq.dogs.tax.table <- as.data.frame(tax_table(phyloseq.dogs))


# Each row of the tax.table is a unique species that is found at least once in the dataset

# You can inspect how the unique phyla by doing the following 
# The "$" operator accesses a column of the data.frame 
unique(phyloseq.dogs.tax.table$phylum)


#you can add the "length" function to give the number of unique phyla 
length(unique(phyloseq.dogs.tax.table$phylum))

```

How many unique family are there?

genera?

```{r}
#| results: hide    
#3) The sample_data function allows us to access the "metadata"
phyloseq.dogs.metadata <- as.data.frame(sample_data(phyloseq.dogs))

#Go through the metadata columns to see what sorts of information we have on the samples 
# Remember, this dataset contains dogs sampled longitudinally 

#Start to think about what might categories might be interesting to test their associations with the microbiome



```

How many unique dogs do we have?

diet groups?

```{r}
#| results: hide    
#We can always subset our phyloseq object to work with a subset of samples if we need to 

# We can do this by using the subset_samples function a given metadata category 

# Let's get a phyloseq object of just puppy samples (indicated by the "host_life_stage column)


phyloseq.dogs.puppies <- subset_samples(phyloseq.dogs, host_life_stage == "puppy")

```

how many puppy samples are there?

# Section 3: Visualization of Microbiome Data using Bar Plots

```{r}
# We will be using a command called "comp_barplot" which takes our phyloseq object and creates relative abundance plots 

#step 1) Use the tax_fix command to make the taxonomy appropriate for the comp_barplot function 
#we will use our 

fixed.physeq <- tax_fix(phyloseq.dogs.puppies)


#Step 2) We can use the comp_barplot to create a barplot 
# It takes two arguments 1) The "fixed" phyloseq object and 2) the taxonomic rank at which to create the plots 
sample_data(fixed.physeq)

comp_barplot(fixed.physeq, tax_level = "species", group_by = "host_age_group", label = NULL)

```
